---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)


Sys.setenv("BIRDDB_HOME"=tempdir())
```

# birddb

<!-- badges: start -->
[![R-CMD-check](https://github.com/cboettig/birddb/workflows/R-CMD-check/badge.svg)](https://github.com/cboettig/birddb/actions)
[![Codecov test coverage](https://codecov.io/gh/cboettig/birddb/branch/main/graph/badge.svg)](https://codecov.io/gh/cboettig/birddb?branch=main)
[![CRAN status](https://www.r-pkg.org/badges/version/birddb)](https://CRAN.R-project.org/package=birddb)
<!-- badges: end -->

The goal of `birddb` is to provide a relational database interface to a local copy of eBird.  `birddb` works by importing the text-based ebird file into a local parquet file using [arrow](https://cran.r-project.org/package=arrow), which can be queried as a relational database using the familiar `dplyr` interface.
`dplyr` translates R-based queries into SQL commands which are past to [`duckdb`](https://duckdb.org), which then queries the parquet database. Unlike the native `arrow` interface, `duckdb` supports the full set of SQL instructions, including windowed operations like `group_by`+`summarise` as well as table joins.


## Installation

You can install the released version of `birddb` from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("birddb")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cboettig/birddb")
```
## Getting Started



```{r}
library(birddb)
library(dplyr)
```

Before you can use `birddb` you will need to download the latest version of eBird from http://ebird.org/ebird/data/download.
Once you have obtained a downloaded copy of the `tar` file, `birddb` can import it for you:
The one-time import of the full data dump is a little slow (about 1 hr in my benchmark) due to the time required to extract the 
tar file and convert the text data into parquet format.

For illustration & testing purposes, we will use the small eBird sample data, included in the package for convenience and testing purposes:

```{r}
tar <- ebird_sample_data()
```

Importing will now create the local parquet-based copies in the default directory given by `ebird_data_dir()`.
Users can set an alternative location by setting the environmental variable `BIRDDB_HOME` to the desired path.

```{r}
import_ebird(tar)
```


Once the data have been downloaded and imported successfully, a user can access the full ebird record quite quickly:

```{r}  
chklists <- checklists()
obs <- observations()

```

Now, we can use `dplyr` to perform standard queries:

```{r}
colnames(obs)
```

```{r}
sp <- obs %>% count(`SCIENTIFIC NAME`, sort=TRUE)

bench::bench_time({collect(sp)})
```



```{r include=FALSE}
Sys.unsetenv("BIRDDB_HOME")

```


