---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

Sys.setenv("BIRDDB_HOME" = tempdir())
```

# birddb

<!-- badges: start -->
[![R-CMD-check](https://github.com/cboettig/birddb/workflows/R-CMD-check/badge.svg)](https://github.com/cboettig/birddb/actions)
[![Codecov test coverage](https://codecov.io/gh/cboettig/birddb/branch/main/graph/badge.svg)](https://codecov.io/gh/cboettig/birddb?branch=main)
[![CRAN status](https://www.r-pkg.org/badges/version/birddb)](https://CRAN.R-project.org/package=birddb)
<!-- badges: end -->

The goal of `birddb` is to provide a relational database interface to a local copy of eBird.  `birddb` works by importing the text-based ebird file into a local parquet file using [arrow](https://cran.r-project.org/package=arrow), which can be queried as a relational database using the familiar `dplyr` interface.
`dplyr` translates R-based queries into SQL commands which are past to [`duckdb`](https://duckdb.org), which then queries the parquet database. Unlike the native `arrow` interface, `duckdb` supports the full set of SQL instructions, including windowed operations like `group_by`+`summarise` as well as table joins.


## Installation

You can install the released version of `birddb` from [CRAN](https://CRAN.R-project.org) (once released):

``` r
install.packages("birddb")
```

And the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("cboettig/birddb")
```

**NOTE**: `birddb` currently requires the dev version of `duckdb`, which you can install using:

```r
install.packages("https://github.com/duckdb/duckdb/releases/download/master-builds/duckdb_r_src.tar.gz", repos = NULL)
```



## Getting Started

```{r message=FALSE}
library(birddb)
library(dplyr)
```

Before you can use `birddb` you will need to download the latest version of the eBird Basic Dataset from http://ebird.org/ebird/data/download.
Once you have obtained a downloaded copy of the `tar` file, `birddb` can import it for you. The one-time import of the full data dump is a little slow (about 1 hr in my benchmark) due to the time required to extract the tar file and convert the text data into parquet format.

For illustration and testing purposes, we will use the small eBird sample data, included in the package for convenience and testing purposes:

```{r}
observations_tar <- birddb::sample_observation_data()
checklists_tar <- birddb::sample_checklist_data()
```

Importing will now create the local parquet-based copies in the default directory given by `ebird_data_dir()`.
Users can set an alternative location by setting the environmental variable `BIRDDB_HOME` to the desired path.

```{r}
import_ebird(observations_tar)
import_ebird(checklists_tar)
```

Once the data have been downloaded and imported successfully, a user can access the full eBird dataset quite quickly:

```{r}  
observations <- observations()
checklists <- checklists()
```

To see the available columns in each dataset use:

```{r}
colnames(observations)
colnames(checklists)
```

Now, we can use `dplyr` to perform standard queries. For example, to see the number of observations for each species in the sample dataset:

```{r}
observations %>% count(scientific_name, sort = TRUE)
```


```{r include=FALSE}
Sys.unsetenv("BIRDDB_HOME")
```
